---
title: "Step1a_environment_setup"
author: "T. Divoll"
date: "2023-09-12"
output: 
  html_document:
  keep_md: TRUE
  df_print: paged
  toc: true
params:
    #don't change these
    datetime: !r (format(Sys.time(), "%Y%m%dT%H:%M:%SZ"))
    user: !r Sys.getenv("LOGNAME")
    data_path: "/oscar/data/tkartzin/projects"
  
    #do change params below this line as needed
    project_code: "test"
---

### Environment Setup
```{r include=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(knitr, here, tidyr, dplyr, readxl, stringr, spgs, filesstrings, lubridate, fs)
here::i_am("./Step1a_env_setup.Rmd")
```

```{r setup, include=FALSE}
# set global chunk parameters here
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
do.call('Sys.setenv', params)
```

#### Set up standard path variables

**Note:** When using the provided test data for training:
Make sure you set the `project_code` param in the YAML to "test" and save as copy of `sample_sheet_test.xlsx` as `sample_sheet.xlsx`!

```{r}
dir.create(here(params$datetime))
work_dir <- file.path(here(params$datetime))

sample_sheet <- read_xlsx(here("sample_sheet.xlsx"), col_types = "text")
```

#### Cleanup on data formats
```{r}
sample_sheet <- sample_sheet %>%
  mutate(across(contains("date"), ~ ymd(.)))

sample_sheet$SampleName <- as.character(sample_sheet$SampleName)
sample_sheet$SampleID <- as.character(sample_sheet$SampleID)
```

The following dates are pulled from the sample sheet and can be used to filter out dates in case of contamination. The unique extraction, PCR, and sequencing dates are cast to lists that we can then index into later.
```{r}
extraction_date <- unique(sample_sheet$`Ext date`)
pcr_date <- unique(sample_sheet$`PCR date`)
sequencing_dates <- unique(sample_sheet$`Seq date`)
folders <- unique(sample_sheet$Folder)
```

Next create a folder for results for each sequencing date in the sample sheet.
```{r}
for (j in seq_along(sequencing_dates)){
  dir.create(file.path(here(work_dir, sequencing_dates[j])))
  dir.create(file.path(here(work_dir, sequencing_dates[j], "results")))
  dir.create(file.path(here(work_dir, sequencing_dates[j], "results/id_data")))
  dir.create(file.path(here(work_dir, sequencing_dates[j], "results/bad_controls_path")))
}

seq_date_dirs <- dir(file.path(here(work_dir)), pattern="^\\d{4}-\\d{2}-\\d{2}", full.names=TRUE)

for (k in seq_along(seq_date_dirs)){
  dir_copy(here("template/src"), seq_date_dirs[k], overwrite=TRUE)
}
```

#### Copy over data and sample sheet
```{r}
samples_path = file.path(params$data_path, params$project_code)

seq_date_names <- data.frame(sample_sheet %>% group_by(Folder) %>% summarise(value = unique(`Seq date`)))

seq_date_names$value <- ymd(seq_date_names$value)

for (m in seq_along(folders)){
  from = file.path(samples_path, folders[m])
  to = file.path(here(work_dir, seq_date_names$value[m]))
  dir_copy(file.path(from, "raw_data"), file.path(to, "renamed_data"), overwrite=TRUE)
}

file.copy(from=here("sample_sheet.xlsx"), to=here(work_dir))

```

#### Generate some derived parameters from the inputs

```{r}
F_primer_reverse <- reverseComplement(params$F_primer)
R_primer_reverse <- reverseComplement(params$R_primer)
F_primer_reverse 
R_primer_reverse
```

```{r}
## make environment variables that can be passed to bash code chunks
Sys.setenv(user=params$user)
Sys.setenv(F_primer=params$F_primer)
Sys.setenv(R_primer=params$R_primer)
Sys.setenv(F_primer_reverse = F_primer_reverse)
Sys.setenv(R_primer_reverse = R_primer_reverse)
```