---
title: "Step 3 - assign taxonomy"
author: "B.Littleford-Colquhoun"
date: "2023-09-21"
output: 
  html_document:
  df_print: paged
  toc: true
---

```{r setup, include=FALSE}
# set global chunk parameters here
knitr::opts_chunk$set(echo = TRUE)
```

## Environment Setup
This notebook can be used to run the OBITools 2 pipeline to assign taxonomy to sequence reads and serves as a tutorial. 

Run through each code chunk and inspect the outputs or select "Run All" from the dropdown above to run the whole pipeline.
```{r include=FALSE}
#install.packages("pacman")
pacman::p_load(here)
here::i_am("src/Step3_taxonomy_assignment.Rmd")
```


#### Set up standard path variables
Here, you want to enter the folder name that you created during Step2
```{r}
DATE<-"YYYY-MM-DD"
FILE_NAME<-"YYYYDDMM_alluniq.clean.fasta"

cleaned_obitools_fasta<-file.path(here(DATE, "results/merged", FILE_NAME))
global_ref_db<-file.path("/oscar/data/tkartzin/global_ref_lib_plants/YYYYMMDD/embl_r143") #-d in ecotag; ecoPCR taxonomy Database name
ref_seq<-file.path("/oscar/data/tkartzin/global_ref_lib_plants/YYYYMMDD/db_P6_r143.fasta") #-R in ecotag; fasta file containing the reference sequences

Sys.setenv(cleaned_obitools_fasta=cleaned_obitools_fasta)
Sys.setenv(global_ref_db=global_ref_db)
Sys.setenv(ref_seq=ref_seq)
Sys.setenv(DATE=DATE)
```


#### Assign taxonomy to each sequence using the global reference library
```{bash}
ecotag -d ${global_ref_db} -R ${ref_seq} ${cleaned_obitools_fasta} > ./results/merged/${DATE}_alluniq.clean.tag.fasta

#Count number of sequences in your file
obicount ./results/merged/${DATE}_alluniq.clean.tag.fasta
```

Un-useful attributes can be removed from sequences
```{bash}
obiannotate --delete-tag=scientific_name_by_db --delete-tag=obiclean_samplecount \
  --delete-tag=obiclean_count --delete-tag=obiclean_singletoncount \
  --delete-tag=obiclean_cluster --delete-tag=obiclean_internalcount \
  --delete-tag=obiclean_head --delete-tag=taxid_by_db --delete-tag=obiclean_headcount \
  --delete-tag=id_status --delete-tag=rank_by_db --delete-tag=order_name \
  --delete-tag=order ./results/merged/${DATE}_alluniq.clean.tag.fasta > \
  ./results/merged/${DATE}_alluniq.clean.tag.ann.fasta

#count number of sequences in file (should match above)
obicount -a ./results/merged/${DATE}_alluniq.clean.tag.ann.fasta
```


#### Create output for use in phyloseq
```{bash}
obitab -o ./results/merged/${DATE}_alluniq.clean.tag.ann.fasta > ./results/merged/${DATE}_alluniq.clean.tag.ann.tab
```

